---
title: "DATA606 Data Project"
author: "Michael Ippolito"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r load-packages, message=FALSE, echo=FALSE}

# Packages needed for map:
# install.packages(c("cowplot", "googleway", "ggplot2", "ggrepel", 
#   "ggspatial", "libwgeom", "sf", "rnaturalearth", "rnaturalearthdata", "rgeos")

# Load libraries
library(tidyverse)
library(psych)
library(RCurl)
library(rvest)
library(kableExtra)
library(GGally)
library(gridExtra)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(rgeos)

```

## DATA606 Data Project {.tabset}

### Abstract

It may be self-evident that the amount of money spent on education by a country will have an impact on the country's literacy rate. But is there a statistically significant relationship? And can that relationship be quantified using a predictive model? It is further presumed that literacy rates will be lower in poorer regions of the world and in those having forms of government in which the head of state holds more power and has less accountability to his or her constituents. This project sought to validate those assumptions and to generate a model that will predict the change in literacy rate based on the studied parameters.

The study confirms that literacy rate is impacted positively by educational spending, with a fairly steep relationship: A one-percent increase in educational spending yields an almost a three-percent rise in literacy rate. There is also evidence to suggest a statistically significant relationship between literacy and both the country’s geographic subregion and its form of government, although it is noted that not all conditions were sufficiently satisfied for the analysis.

### Part 1 - Introduction

It may be self-evident that the amount of money spent on education by a country will have an impact on the country's literacy rate. But is there a statistically significant relationship? And can that relationship be quantified using a predictive model?

It is further presumed that literacy rates will be lower in poorer regions of the world and in those having forms of government in which the head of state holds more power and has less accountability to his or her constituents. This project seeks to validate those assumptions and to generate a model that will predict the change in literacy rate based on the studied parameters.


### Part 2 - Data

The data came from several sources:

* *Regional data* is from a public dataset on Github$^1$.

* *Government spending on education* is posted on worldbank.org and is collected annually from the UNESCO Institute for Statistics; data from the years 1960 to 2020 was available for download$^2$.

* *Literacy rates* were obtained from an aggregated dataset on wikimedia.org; the data was originally collected by worldbank.org, the CIA World Factbook, and other sources. The dataset includes some historical data as old as 1475 but with a majority of data points between 1960 and 2015$^3$. For the purposes of this study, only data from 1960 or newer was used.

* *Forms of government* were taken from Wikipedia$^4$.

Data preparation included a few tidying steps:

* Removing footnotes and general tidying of the data
* Standardizing country codes across data sets
* Standardizing field names

```{r data_prep}

##########################
# Regions
##########################

# Load countries by region
# from https://raw.githubusercontent.com/lukes/ISO-3166-Countries-with-Regional-Codes/master/all/all.csv
regions_full <- read.csv("https://raw.githubusercontent.com/mmippolito/cuny/main/data606/project/country_regions.csv")

# Standarize field names
regions <- regions_full %>%
  rename(
    "country_numeric_code" = "country.code",
    "country_code" = "alpha.3",
    "subregion" = "sub.region"
  )


##########################
# Government spending
##########################

# Load govt spending on education data set
# from https://data.worldbank.org/indicator/SE.XPD.TOTL.GD.ZS
edspending_full <- read.csv("https://raw.githubusercontent.com/mmippolito/cuny/main/data606/project/govt_ed_spending.csv")

# Standardize field names
edspending <- edspending_full %>%
  rename(
    "country" = "Country.Name",
    "country_code" = "Country.Code"
  ) %>%
  select(-`Indicator.Name`, -`Indicator.Code`, -`X`)

# Remove Xs from years in column names
edspending <- rename_with(edspending, function(x) ifelse(substr(x, 1, 1) == "X", substr(x, 2, 5), x))

# Gather columns
edspending <- edspending %>% gather(3:63, key = "year", value = "pct_gdp")

# Convert years to numeric
edspending$year <- as.numeric(edspending$year)


##########################
# Literacy rates
##########################

# Load literacy rates
# from https://commons.wikimedia.org/wiki/Data:Cross-country_literacy_rates_-_World_Bank,_CIA_World_Factbook,_and_other_sources_(OWID_2762).tab
lit_full <- read.csv("https://raw.githubusercontent.com/mmippolito/cuny/main/data606/project/cross-country-literacy-rates.csv")

# Standardize field names
lit <- lit_full %>%
    rename(
        "country" = "Entity",
        "country_code" = "Code",
        "year" = "Year",
        "lit_rate" = "Literacy.rates..World.Bank..CIA.World.Factbook..and.other.sources."
    ) %>%
    filter(year >= 1960)


##########################
# Forms of government
##########################

# Load forms of government from CSV if it exists, otherwise parse from Wikipedia
# from https://en.wikipedia.org/wiki/List_of_countries_by_system_of_government
if (url.exists("https://raw.githubusercontent.com/mmippolito/cuny/main/data606/project/govt_forms.csv")) {
    govts <- read.csv("https://raw.githubusercontent.com/mmippolito/cuny/main/data606/project/govt_forms.csv", header = T)
} else {
    # Scrape and parse tables
    govts_html <- read_html("https://en.wikipedia.org/wiki/List_of_countries_by_system_of_government", encoding = "UTF-8")
    govts_tables <- html_table(govts_html, fill = TRUE)
    govts <- data.frame(govts_tables[6])
    # Standarize field names
    govts <- govts %>%
        rename(
            country = Name, 
            govt = Constitutional.form, 
            head_of_state = Head.of.state, 
            exec_legitimacy = Basis.of.executive.legitimacy)
    # Remove footnotes
    govts$exec_legitimacy <- str_replace_all(govts$exec_legitimacy, '\\[.+?\\]', '')
    # Change provisional gov't with no head to "no" instead of "n/a"
    govts$head_of_state <- ifelse(govts$head_of_state == "n/a", "no", govts$head_of_state)
    # Lower case
    govts$govt <- tolower(govts$govt)
    govts$head_of_state <- tolower(govts$head_of_state)
    govts$exec_legitimacy <- tolower(govts$exec_legitimacy)
    # Standardize names
    govts <- govts %>%
        mutate(country = case_when(
            country == "Bahamas, The" ~ "Bahamas",
            country == "China, People's Republic of" ~ "China",
            country == "Congo, Democratic Republic of the" ~ "Democratic Republic of Congo",
            country == "Congo, Republic of the" ~ "Congo",
            country == "Côte d'Ivoire" ~ "Cote d'Ivoire",
            country == "Czech Republic" ~ "Czechia",
            country == "East Timor" ~ "Timor",
            country == "Federated States of Micronesia" ~ "Micronesia",
            country == "Gambia, The" ~ "Gambia",
            country == "Korea, North" ~ "North Korea",
            country == "Korea, South" ~ "South Korea",
            country == "São Tomé and Príncipe" ~ "Sao Tome and Principe",
            country == "Vatican City" ~ "Vatican",
            TRUE ~ country)
        )
    # Add country codes by joining to lit_sum
    govts <-govts %>% 
        left_join(
            lit %>% group_by(country, country_code) %>% summarize(n(), .groups = "keep")
            , by = c("country")) %>%
        select(country, govt, head_of_state, exec_legitimacy, country_code)
    # Save to CSV
    # Create short form of exec_legitimacy
    govts <- govts %>%
        mutate(exec_legit_short = case_when(
            exec_legitimacy == "all authority vested in absolute monarch" ~ 
                "having absolute authority",
            exec_legitimacy == "ministry is subject to parliamentary confidence" ~ 
                "accountable to legislature",
            exec_legitimacy == "monarch personally exercises power in concert with other institutions" ~ 
                "sharing power",
            exec_legitimacy == "no constitutionally-defined basis to current regime" ~ 
                "with no contsitutional legitimacy",
            exec_legitimacy == "power constitutionally linked to a single political movement" ~ 
                "constitutionally granted power by single party",
            exec_legitimacy == "presidency independent of legislature; ministry is subject to parliamentary confidence" ~ 
                "independent of and accountable to legislature",
            exec_legitimacy == "presidency is elected by legislature; ministry may be, or not be, subject to parliamentary confidence" ~ 
                "elected by legislature",
            exec_legitimacy == "presidency is independent of legislature" ~ 
                "independent of legislature",
            TRUE ~ '')
        )
    govts$govt_type_short <- paste0(govts$govt, " with ", govts$head_of_state, " head\n", govts$exec_legit_short)
}

```

Because the data came from disparate sources, there were gaps that had to be addressed before proceeding with analysis. For example, many countries had sparser data on literacy rates compared to government spending on education:

```{r data_gap1, ECHO=FALSE}

regions %>%
    select(country_code) %>%
    left_join(edspending, by = c("country_code")) %>%
    group_by(country_code) %>%
    summarize(edspending_year_count = n()) %>%
    full_join(lit, by = c("country_code")) %>%
    group_by(country_code, edspending_year_count) %>%
    summarize(litrate_year_count = n(), .groups = "keep") %>%
    full_join(govts, by = c("country_code")) %>%
    group_by(country_code, edspending_year_count, litrate_year_count) %>%
    summarize(govt_count = n(), .groups = "keep") %>%
    group_by() %>%
    summarize(countries = n(), ed_spending_year_count = sum(edspending_year_count, na.rm = T),
        litrate_year_count = sum(litrate_year_count, na.rm = T), govt_count = sum(govt_count, na.rm = T)) %>%
    kable(caption = "<i><font color=#000000><b>Table 1.</b> Summary of observations</font></i>") %>% 
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 13)

```

In addition, some countries had no country code listed, while others had no literacy rate or educational spending data:

```{r data_gap2, ECHO=FALSE}

regions %>%
    select(country_code) %>%
    left_join(edspending, by = c("country_code")) %>%
    group_by(country_code) %>%
    summarize(edspending_year_count = n()) %>%
    full_join(lit, by = c("country_code")) %>%
    group_by(country_code, edspending_year_count) %>%
    summarize(litrate_year_count = n(), .groups = "keep") %>%
    full_join(govts, by = c("country_code")) %>%
    group_by(country_code, edspending_year_count, litrate_year_count) %>%
    summarize(govt_count = n(), .groups = "keep") %>%
    arrange(country_code) %>%
    filter(is.na(edspending_year_count) | is.na(litrate_year_count) | is.na(govt_count)) %>%
    kable(caption = "<i><font color=#000000><b>Table 1.5.</b> Data gaps</font></i>") %>% 
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 13)
    
```

For these reasons, the data couldn't be reliably compared on a year-by-year basis. Instead, the data was taken as an aggregate, using mean values for literacy rate and educational spending per country.

```{r data_agg}

##########################
# Data aggregation
##########################

# Calculate mean literacy rate per country
lit_sum <- lit %>% 
    group_by(country, country_code) %>%
    summarize(lit_rate = mean(lit_rate, na.rm = T), .groups = "keep")

# Calculate mean education spending per country
ed_sum <- edspending %>% group_by(country, country_code) %>%
    summarize(pct_gdp = mean(pct_gdp, na.rm = T), .groups = "keep")

# Join the summarized tables
j <- lit_sum %>% 
    full_join(ed_sum, by = c("country_code")) %>%
    full_join(govts, by = c("country_code")) %>%
    left_join(regions, by = c("country_code"))

# Filter out rows with no country code or subregion; these are regional aggregations
j <- j %>%
    filter(str_length(country_code) > 0 & str_length(region) > 0)

# Filter out NaNs
#j <- j %>%
#  filter(!is.na(pct_gdp)) %>%
#  filter(!is.na(lit_rate))

# Select only relevant fields
j <- j %>%
    select(country_code, country = country.x, region, subregion, 
        pct_gdp, lit_rate, govt, head_of_state, exec_legitimacy, exec_legit_short, govt_type_short)

# Add rankings
j$rank_pct_gdp <- rank(desc(j$pct_gdp))
j$rank_lit_rate <- rank(desc(j$lit_rate))

```


### Part 3 - Exploratory data analysis

Data overview:

```{r explore1}

# Data overview
j %>%
    arrange(country_code) %>%
    kable(caption = "<i><font color=#000000><b>Table 2.</b> Sample data</font></i>") %>% 
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 13)

```

Literacy rates (dependent variable):

```{r explore2}

# Summary statistics for literacy rate
describe(j$lit_rate) %>%
    kable(caption = "<i><font color=#000000><b>Table 3.</b> Summary stats - literacy rate</font></i>") %>% 
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 13)

# Histogram of literacy rate
j %>%
    drop_na(lit_rate) %>%
    ggplot() +
    geom_histogram(aes(x = lit_rate), bins = 25, fill = 'lightblue') +
    xlab("Literacy rate") + ylab("Countries") +
    ggtitle("Figure 1. Literacy rate") + 
    theme_light()

```

Educational spending (independent, numerical variable):

```{r explore3}

# Summary statistics for percent GDP education spending
describe(j$pct_gdp) %>%
    kable(caption = "<i><font color=#000000><b>Table 4.</b> Summary stats - educational spending</font></i>") %>% 
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 13)

# Histogram of GDP education spending
j %>%
    drop_na(pct_gdp) %>%
    ggplot() +
    geom_histogram(aes(x = pct_gdp), bins = 20, fill = 'lightblue') +
    xlab("% GDP") + ylab("Countries") +
    ggtitle("Figure 2. Percentage of GDP Spent on Education") +
    theme_light()

```

Regions (independent, categorical variable):

```{r explore4}

# Regions table
regions %>% 
    group_by(region, subregion) %>%
    summarize(n = n(), .groups = "keep") %>%
    select(-n) %>% 
    arrange(region, subregion) %>%
    kable(caption = "<i><font color=#000000><b>Table 4.5.</b> Regions</font></i>") %>% 
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 13)

# Bar chart of regions
j %>%
    drop_na(subregion) %>%
    group_by(subregion) %>%
    summarize(n = n()) %>%
    ggplot() +
    geom_bar(aes(x = reorder(subregion, n), y = n), fill = 'lightblue', stat = "identity") +
    xlab("Subregion") + ylab("Countries") +
    ggtitle("Figure 3. Subregions") +
    theme_light() +
    coord_flip()

```

Forms of government (independent, categorical variable):

```{r explore5}

# Governments table
govts %>% 
    group_by(govt, head_of_state, exec_legitimacy) %>%
    summarize(num_countries = n(), .groups = "keep") %>%
    arrange(desc(num_countries)) %>%
    kable(caption = "<i><font color=#000000><b>Table 4.7.</b> Forms of government</font></i>") %>% 
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 13)

# Bar chart of governments
govts %>%
    group_by(govt_type_short) %>%
    summarize(n = n(), .groups = "keep") %>%
    ggplot() +
    geom_bar(aes(x = reorder(govt_type_short, n), y = n), fill = 'lightblue', stat = "identity") +
    xlab("Form of goverment") + ylab("Countries") +
    ggtitle("Figure 4. Forms of government") +
    theme_light() +
    coord_flip()

```

Explore relationships between explanatory and response variables:

```{r explore6}

# Scatter plot of literacy rate vs education spending
j %>% 
    drop_na(pct_gdp, lit_rate) %>%
    ggplot() +
    geom_point(aes(x = pct_gdp, y = lit_rate)) +
    theme_light() +
    xlab("Percent GDP") +
    ylab("Literacy rate") +
    ggtitle("Figure 5. Literacy rate vs educational spending")

# Box plots of literacy rate by region
j %>%
    drop_na(lit_rate) %>%
    ggplot() +
    geom_boxplot(aes(x = reorder(subregion, -lit_rate), y = lit_rate)) +
    coord_flip() +
    theme_light() +
    xlab("Subregion") +
    ylab("Literacy rate") +
    ggtitle("Figure 6. Literacy rate by subregion")

# Box plots of literacy rate by form of government
j %>%
    drop_na(lit_rate) %>%
    drop_na(govt_type_short) %>%
    ggplot() +
    geom_boxplot(aes(x = reorder(govt_type_short, -lit_rate), y = lit_rate)) +
    xlab("Form of goverment") + ylab("Literacy rate") +
    ggtitle("Figure 7. Literacy rate by form of government") +
    theme_light() +
    coord_flip()

# Set world class for maps
world <- ne_countries(scale = "medium", returnclass = "sf", type = "countries")

# Map of literacy rate
world %>%
    inner_join(j, by = c("iso_a3" = "country_code")) %>%
    drop_na(lit_rate) %>%
    ggplot() +
    geom_sf(aes(fill = lit_rate)) +
    scale_fill_viridis_c(option = "cividis", trans = "sqrt", direction = -1) +
    ggtitle("Figure 8. Literacy rate") +
    labs(fill = "Literacy rate")

```

Explore independence of explanatory variables:

```{r explore7}

# Box plots of ed spending by subregion
j %>%
    drop_na(pct_gdp) %>%
    ggplot() +
    geom_boxplot(aes(x = reorder(subregion, -pct_gdp), y = pct_gdp)) +
    coord_flip() +
    theme_light() +
    xlab("Subregion") +
    ylab("% GDP") +
    ggtitle("Figure 9. Educational spending by subregion")

# Box plots of ed spending by form of government
j %>%
    drop_na(pct_gdp, govt_type_short) %>%
    ggplot() +
    geom_boxplot(aes(x = reorder(govt_type_short, -pct_gdp), y = pct_gdp)) +
    coord_flip() +
    theme_light() +
    xlab("Form of government") +
    ylab("% GDP") +
    ggtitle("Figure 10. Educational spending by form of government")

# Bar plot of form of government by region
j %>%
    drop_na(govt, region) %>%
    ggplot() +
    geom_bar(aes(x = region, fill = govt), position = "dodge") +
    coord_flip() +
    theme_light() +
    xlab("Region") +
    ylab("Countries") +
    ggtitle("Figure 11. Form of government by region")

# Map of subregions
world %>%
    inner_join(j, by = c("iso_a3" = "country_code"))  %>%
    ggplot() +
    geom_sf(aes(fill = subregion.y)) +
    ggtitle("Figure 12. Subregions of the world") +
    labs(fill = "Subregion")

# Map of forms of government
world %>%
    inner_join(govts, by = c("iso_a3" = "country_code")) %>%
    ggplot() +
    geom_sf(aes(fill = govt)) +
    scale_fill_brewer(palette="PuBu") +
    ggtitle("Figure 13. Forms of government") +
    labs(fill = "Government")

# Map of educational spending
world %>%
    inner_join(j, by = c("iso_a3" = "country_code")) %>%
    drop_na(pct_gdp) %>%
    ggplot() +
    geom_sf(aes(fill = pct_gdp)) +
    scale_fill_viridis_c(option = "cividis", trans = "sqrt", direction = -1) +
    ggtitle("Figure 14. Educational spending") +
    labs(fill = "% GDP")

```


### Part 4 - Inference

One goal of the project was to evaluate whether there is a statistically significant difference in literacy rate and educational spending across regions as well as across government types. To evaluate this, a series of ANOVA analyses were performed. The null hypothesis for each analysis was that there was no difference in rate for the variable in question; for example, for the first test, $H_0$ = there is no difference in educational spending rate among subregions. The alternative hypothesis would be the opposite: $H_A$ = there is a statistically significant difference in educational spending rate among subregions. Similar comparisons were made for the remaining ANOVA tests.

It is noted that not all of the conditions for ANOVA were met. Independence was assumed, but while education spending was fairly normal across subregions and government types, the same could not be said of literacy rate, which was strongly left-skewed. Similarly, the box plots also show that variance was not constant for literacy rates, while it was more constant for educational spending.

```{r inf1}

# Literacy rates by subregion
region_summ <- j %>% 
    group_by(region, subregion) %>%
    summarize(
        n_pct_gdp = n(),
        sd_pct_gdp = sd(pct_gdp, na.rm = T),
        mean_pct_gdp = mean(pct_gdp, na.rm = T),
        n_lit_rate = n(),
        sd_lit_rate = sd(lit_rate, na.rm = T),
        mean_lit_rate = mean(lit_rate, na.rm = T),
        .groups = "keep"
    )

# Literacy rates by govt
govt_summ <- j %>% 
    drop_na(govt_type_short) %>%
    group_by(govt_type_short) %>%
    summarize(
        n_pct_gdp = n(),
        sd_pct_gdp = sd(pct_gdp, na.rm = T),
        mean_pct_gdp = mean(pct_gdp, na.rm = T),
        n_lit_rate = n(),
        sd_lit_rate = sd(lit_rate, na.rm = T),
        mean_lit_rate = mean(lit_rate, na.rm = T),
        .groups = "keep"
    ) %>%
    mutate(sd_pct_gdp = ifelse(is.na(sd_pct_gdp), 0, sd_pct_gdp)) %>%
    mutate(sd_lit_rate = ifelse(is.na(sd_lit_rate), 0, sd_lit_rate))

# Find overall rate - ed spending
mean_pct_gdp = mean(j$pct_gdp, na.rm = T)
sd_pct_gdp = sd(j$pct_gdp, na.rm = T)
print(paste0("Mean spending on education across all countries = ", round(mean_pct_gdp, 3), "%, sd = ", round(sd_pct_gdp, 3)))

# Find overall rate - literacy rate
mean_lit_rate = mean(j$lit_rate, na.rm = T)
sd_lit_rate = sd(j$lit_rate, na.rm = T)
print(paste0("Mean literacy rate across all countries = ", round(mean_lit_rate, 3), "%, sd = ", round(sd_lit_rate, 3)))
print("")

# Histograms - ed spending by subregion
j %>%
    drop_na(pct_gdp) %>%
    ggplot(aes(x = pct_gdp)) +
    geom_histogram(binwidth = 0.5) +
    facet_wrap(~subregion, ncol = 3) + 
    ggtitle("Figure 15. Educational spending by subregion") +
    xlab("Educational spending (% GDP)") + ylab("")

# Histograms - literacy rate by subregion
j %>%
    drop_na(lit_rate) %>%
    ggplot(aes(x = lit_rate)) +
    geom_histogram(binwidth = 7) +
    facet_wrap(~subregion, ncol = 3) + 
    ggtitle("Figure 16. Literacy rate by subregion") +
    xlab("Literacy rate (%)") + ylab("")

# Histograms - ed spending by govt
j %>%
    drop_na(pct_gdp) %>%
    ggplot(aes(x = pct_gdp)) +
    geom_histogram(binwidth = 0.5) +
    facet_wrap(~govt_type_short, ncol = 3) + 
    ggtitle("Figure 17. Educational spending by government type") +
    xlab("Educational spending (% GDP)") + ylab("")

# Histograms - literacy rate by govt
j %>%
    drop_na(lit_rate) %>%
    ggplot(aes(x = lit_rate)) +
    geom_histogram(binwidth = 7) +
    facet_wrap(~govt_type_short, ncol = 3) + 
    ggtitle("Figure 18. Literacy rate by government type") +
    xlab("Literacy rate (%)") + ylab("")

# ANOVA - ed spending vs subregion
aov_ed_region <- aov(pct_gdp ~ subregion, data = j)
print(paste0("ANOVA results - educational spending vs subregion"))
summary(aov_ed_region)

# ANOVA - ed spending vs govt
aov_ed_govt <- aov(pct_gdp ~ govt_type_short, data = j)
print(paste0("ANOVA results - educational spending vs government type"))
summary(aov_ed_govt)

# ANOVA - lit rate vs subregion
aov_lit_region <- aov(lit_rate ~ subregion, data = j)
print(paste0("ANOVA results - literacy rate vs subregion"))
summary(aov_lit_region)

# ANOVA - lit rate vs govt
aov_lit_govt <- aov(lit_rate ~ govt_type_short, data = j)
print(paste0("ANOVA results - literacy rate vs government type"))
summary(aov_lit_govt)

```

Next, a linear model was fit to the literacy rate vs ed spending data. It should be noted that not all of the conditions for least squares are *not* met:

1. Linearity: The data appear to be fairly linear, with a positive relationship.
2. Nearly normal residuals: As shown in the histogram and QQ plot, the residuals are significantly skewed to the left.
3. Constant variability: Variability isn't constant, but narrows for higher values.
4. Independent observations: The observations are independent.

A second model was fit that ignored high-leverage points (percent gdp > 7.75%). Contrary to what I expected, the line didn't change much.

```{r lm}

# Linear model - literacy rate vs ed spending
lit_lm <- lm(lit_rate ~ pct_gdp, data = j, na.action = na.omit)
summary(lit_lm)

# Line of best fit
j %>%
    drop_na(pct_gdp, lit_rate) %>%
    ggplot(aes(x = pct_gdp, y = lit_rate)) +
    geom_point() + 
    geom_smooth(formula = y ~ x, method = "lm", se = T) +
    coord_cartesian(ylim = c(0, 100)) +
    xlab("Government spending (% GDP)") + ylab("Literacy rate (%)") +
    ggtitle("Figure 19. Literacy rate vs educational spending")

# Residuals plot
lit_lm %>%
    ggplot(aes(x = .fitted, y = .resid)) + 
    geom_point() +
    geom_hline(yintercept = 0, linetype = "dashed") +
    xlab("Fitted values") + ylab("Residuals") +
    ggtitle("Figure 20. Fitted values vs residuals")

# Residuals histogram
lit_lm %>%
    ggplot(aes(x = .resid)) +
    geom_histogram(binwidth = 4) +
    xlab("Residuals") + ylab("Count") +
    ggtitle("Figure 21. Residuals histogram")

# Probability plot
lit_lm %>%
    ggplot(aes(sample = .resid)) +
    stat_qq() +
    xlab("Theoretical values") + ylab("Actual values") +
    ggtitle("Figure 22. Probability plot")

# Removing high-leverage/influential points
lit_lm_2 <- lm(lit_rate ~ pct_gdp, data = filter(j, pct_gdp < 7.75), na.action = na.omit)
summary(lit_lm_2)
j %>%
    drop_na(pct_gdp, lit_rate) %>%
    filter(pct_gdp < 7.75) %>%
    ggplot(aes(x = pct_gdp, y = lit_rate)) +
    geom_point() + 
    geom_smooth(formula = y ~ x, method = "lm", se = T) +
    coord_cartesian(ylim = c(0, 100)) +
    xlab("Fitted values") + ylab("Residuals") +
    ggtitle("Figure 22.5. Literacy rate vs educational spending (high-leverage values removed)")

```

One additional goal of the project was to evaluate whether adding the two categorical variables (subregion and government type) would influence the linear model. First, a backward elimination by p-value strategy was used to eliminate variables. But that resulted in removing the pct_gdp variable and ending up with two categorical variables having p-values greater than 0.05. So a different strategy was employed using forward selection.

```{r mutli_fwd_sel}

# Prepare variables
incl_pct_gdp <- c('include', '', '', 'include', 'include', 'include')
incl_subregion <- c('', 'include', '', 'include', '', 'include')
incl_govt_type <- c('', '', 'include', '', 'include', 'include')
adjrsq <- c()

# Fit linear model - pct_gdp
lm_mult <- lm(lit_rate ~ pct_gdp, data = j)
adjrsq <- c(adjrsq, summary(lm_mult)$adj.r.squared)
summary(lm_mult)
# adj r-squared = 0.07742

# subregion
lm_mult <- lm(lit_rate ~ subregion, data = j)
adjrsq <- c(adjrsq, summary(lm_mult)$adj.r.squared)
summary(lm_mult)
# adj r-squared = 0.5086

# govt type
lm_mult <- lm(lit_rate ~ govt_type_short, data = j)
adjrsq <- c(adjrsq, summary(lm_mult)$adj.r.squared)
summary(lm_mult)
# adj r-squared = 0.1334

# Now start with pct_gdp type and add in subregion
lm_mult <- lm(lit_rate ~ pct_gdp + subregion, data = j)
adjrsq <- c(adjrsq, summary(lm_mult)$adj.r.squared)
summary(lm_mult)
# adj r-squared = 0.5616 (higher than pct_gdp alone)

# pct_gdp type + govt_type
lm_mult <- lm(lit_rate ~ pct_gdp + govt_type_short, data = j)
adjrsq <- c(adjrsq, summary(lm_mult)$adj.r.squared)
summary(lm_mult)
# adj r-squared = 0.1724 (lower than pct_gdp alone)

# pct_gdp type + subregion + govt_type
lm_mult <- lm(lit_rate ~ pct_gdp + subregion + govt_type_short, data = j)
adjrsq <- c(adjrsq, summary(lm_mult)$adj.r.squared)
summary(lm_mult)
# adj r-squared = 0.5546 (lower than pct_gdp and subregion)

# data frame
df_models <- data.frame(adj_rsquared = adjrsq, pct_gdp = incl_pct_gdp, 
    subregion = incl_subregion, govt_type = incl_govt_type)
df_models %>%
    kable(caption = "<i><font color=#000000><b>Table 5.</b> Linear model results</font></i>") %>% 
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 13)

# Best model:
lm_mult <- lm(lit_rate ~ pct_gdp + subregion, data = j)
summary(lm_mult)

```

Using forward selection, the best model is from using educational spending by itself, without subregion or government type as additional predictors.

### Part 5 - Conclusion

Based on the results of this analysis, literacy rate is confirmed to be impacted positively by educational spending. It is noted that the impact is fairly steep, with every percentage point of GDP spent on education having a three-percent rise in literacy rate.

The results of the ANOVA analysis indicate strongly that there is a statistically significant difference between both literacy rate and educational spending among regions and, additionally, among various government types. Not surprisingly, higher literacy rates are enjoyed by republics having an elected head of state who is accountable to his or her constituents, while lower literacy rates were observed in countries with provisional governments. While it was expected that literacy rates in sub-Saharan Africa would be among the lowest, it was somewhat surprising that Central Asia boasted the highest overall literacy rates. These countries include the former Soviet republics of Kazakhstan, Kyrgyzstan, Tajikistan, Turkmenistan, Uzbekistan. And while their overall literacy rate is, indeed, very high, according to the UN's definition of literacy, many people in these countries are "functionally" illiterate.$^5$ (UNESCO defines literacy as "the ability to use reading, writing and numeracy skills for effective functioning and development of the individual and the community. A person is literate who can, with understanding, both read and write a short statement on his or her everyday life."$^6$ Conversely, "a person is functionally literate who can engage in all those activities in which literacy is required for effective functioning of his group and community and also for enabling him to continue to use reading, writing, and calculation for his own and the community’s development."$^7$) Another interesting outlier is the amount of money spent on education by Micronesia. While its average 7.9% of GDP spent on education, it should be noted that the U.S. funds around 90% of that amount in exchange for access to the islands for military purposes.$^8$

As noted above, not all conditions for ANOVA were met: Literacy rate data were significantly left-skewed, and the variance was not constant. Similarly, not all conditions for fitting a linear model were met: The residuals were skewed to the left, and variability narrowed at higher values.

In conclusion, there was significant statistical evidence to suggest literacy rate is also affected by subregion of the world and by the type of government in place. Perhaps more significantly, there was strong correlation between government spending on education (as a percentage of GDP) and literacy rate, with a three-fold increase in literacy for every percentage point increase in education spending.

### References

$^1$ https://github.com/lukes/ISO-3166-Countries-with-Regional-Codes/blob/master/all/all.csv
$^2$ https://data.worldbank.org/indicator/SE.XPD.TOTL.GD.ZS
$^3$ https://commons.wikimedia.org/wiki/Data:Cross-country_literacy_rates_-_World_Bank,_CIA_World_Factbook,_and_other_sources_(OWID_2762).tab
$^4$ https://en.wikipedia.org/wiki/List_of_countries_by_system_of_government
$^5$ https://www.dvv-international.de/en/adult-education-and-development/editions/aed-662006/education-for-all-and-literacy/review-of-central-asian-countries
$^6$ https://www.fipfoundation.org/pictograms-support/about-health-literacy/#:~:text=UNESCO%20provides%20a%20definition%20of,his%20or%20her%20everyday%20life.%E2%80%9D
$^7$ https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5102880/
$^8$ https://www.grassrootinstitute.org/2016/09/us-policies-may-hurt-micronesians-more-than-they-help/
