---
title: DATA606 Data Project
author: Michael Ippolito
---

```{r load-packages, message=FALSE, echo=FALSE}

# Packages needed for map:
# install.packages(c("cowplot", "googleway", "ggplot2", "ggrepel", 
#   "ggspatial", "libwgeom", "sf", "rnaturalearth", "rnaturalearthdata", "rgeos")

# Load libraries
library(tidyverse)
library(psych)
library(RCurl)
library(rvest)
library(kableExtra)
library(GGally)
library(gridExtra)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(rgeos)

```

## DATA606 Data Project {.tabset}

### Abstract

The...

### Part 1 - Introduction

It may be self-evident that the amount of money spent on education by a country will have an impact on the country's literacy rate. But is there a statistically significant relationship? And can that relationship be quantified using a predictive model?

It is further presumed that literacy rates will be lower in poorer regions of the world and in those having forms of government in which the head of state holds more power and has less accountability to his or her constituents. This project seeks to validate those assumptions and to generate a model that will predict the change in literacy rate based on the studied parameters.


### Part 2 - Data

The data came from several sources:

* *Regional data* is from a public dataset on Github$^1$.

* *Government spending on education* is posted on worldbank.org and is collected annually from the UNESCO Institute for Statistics; data from the years 1960 to 2020 was available for download$^2$.

* *Literacy rates* were obtained from an aggregated dataset on wikimedia.org; the data was originally collected by worldbank.org, the CIA World Factbook, and other sources. The dataset includes some historical data as old as 1475 but with a majority of data points between 1960 and 2015$^3$.

* *Forms of government* were taken from Wikipedia$^4$.

Data preparation included a few tidying steps:

* Removing footnotes and general tidying of the data
* Standardizing country codes across data sets
* Standardizing field names

```{r data_prep}

##########################
# Regions
##########################

# Load countries by region
# from https://raw.githubusercontent.com/lukes/ISO-3166-Countries-with-Regional-Codes/master/all/all.csv
regions_full <- read.csv("https://raw.githubusercontent.com/mmippolito/cuny/main/data606/project/country_regions.csv")

# Standarize field names
regions <- regions_full %>%
  rename(
    "country_numeric_code" = "country.code",
    "country_code" = "alpha.3",
    "subregion" = "sub.region"
  )


##########################
# Government spending
##########################

# Load govt spending on education data set
# from https://data.worldbank.org/indicator/SE.XPD.TOTL.GD.ZS
edspending_full <- read.csv("https://raw.githubusercontent.com/mmippolito/cuny/main/data606/project/govt_ed_spending.csv")

# Standardize field names
edspending <- edspending_full %>%
  rename(
    "country" = "Country.Name",
    "country_code" = "Country.Code"
  ) %>%
  select(-`Indicator.Name`, -`Indicator.Code`, -`X`)

# Remove Xs from years in column names
edspending <- rename_with(edspending, function(x) ifelse(substr(x, 1, 1) == "X", substr(x, 2, 5), x))

# Gather columns
edspending <- edspending %>% gather(3:63, key = "year", value = "pct_gdp")

# Convert years to numeric
edspending$year <- as.numeric(edspending$year)


##########################
# Literacy rates
##########################

# Load literacy rates
# from https://commons.wikimedia.org/wiki/Data:Cross-country_literacy_rates_-_World_Bank,_CIA_World_Factbook,_and_other_sources_(OWID_2762).tab
lit_full <- read.csv("https://raw.githubusercontent.com/mmippolito/cuny/main/data606/project/cross-country-literacy-rates.csv")

# Standardize field names
lit <- lit_full %>%
  rename(
    "country" = "Entity",
    "country_code" = "Code",
    "year" = "Year",
    "lit_rate" = "Literacy.rates..World.Bank..CIA.World.Factbook..and.other.sources."
  )


##########################
# Forms of government
##########################

# Load forms of government from CSV if it exists, otherwise parse from Wikipedia
# from https://en.wikipedia.org/wiki/List_of_countries_by_system_of_government
if (url.exists("https://raw.githubusercontent.com/mmippolito/cuny/main/data606/project/govt_forms.csv")) {
    govts <- read.csv("https://raw.githubusercontent.com/mmippolito/cuny/main/data606/project/govt_forms.csv", header = T)
} else {
    # Scrape and parse tables
    govts_html <- read_html("https://en.wikipedia.org/wiki/List_of_countries_by_system_of_government", encoding = "UTF-8")
    govts_tables <- html_table(govts_html, fill = TRUE)
    govts <- data.frame(govts_tables[6])
    # Standarize field names
    govts <- govts %>%
        rename(
            country = Name, 
            govt = Constitutional.form, 
            head_of_state = Head.of.state, 
            exec_legitimacy = Basis.of.executive.legitimacy)
    # Remove footnotes
    govts$exec_legitimacy <- str_replace_all(govts$exec_legitimacy, '\\[.+?\\]', '')
    # Change provisional gov't with no head to "no" instead of "n/a"
    govts$head_of_state <- ifelse(govts$head_of_state == "n/a", "no", govts$head_of_state)
    # Lower case
    govts$govt <- tolower(govts$govt)
    govts$head_of_state <- tolower(govts$head_of_state)
    govts$exec_legitimacy <- tolower(govts$exec_legitimacy)
    # Standardize names
    govts <- govts %>%
        mutate(country = case_when(
            country == "Bahamas, The" ~ "Bahamas",
            country == "China, People's Republic of" ~ "China",
            country == "Congo, Democratic Republic of the" ~ "Democratic Republic of Congo",
            country == "Congo, Republic of the" ~ "Congo",
            country == "Côte d'Ivoire" ~ "Cote d'Ivoire",
            country == "Czech Republic" ~ "Czechia",
            country == "East Timor" ~ "Timor",
            country == "Federated States of Micronesia" ~ "Micronesia",
            country == "Gambia, The" ~ "Gambia",
            country == "Korea, North" ~ "North Korea",
            country == "Korea, South" ~ "South Korea",
            country == "São Tomé and Príncipe" ~ "Sao Tome and Principe",
            country == "Vatican City" ~ "Vatican",
            TRUE ~ country)
        )
    # Add country codes by joining to lit_sum
    govts <-govts %>% 
        left_join(
            lit %>% group_by(country, country_code) %>% summarize(n(), .groups = "keep")
            , by = c("country")) %>%
        select(country, govt, head_of_state, exec_legitimacy, country_code)
    # Save to CSV
    # Create short form of exec_legitimacy
    govts <- govts %>%
        mutate(exec_legit_short = case_when(
            exec_legitimacy == "all authority vested in absolute monarch" ~ 
                "having absolute authority",
            exec_legitimacy == "ministry is subject to parliamentary confidence" ~ 
                "accountable to legislature",
            exec_legitimacy == "monarch personally exercises power in concert with other institutions" ~ 
                "sharing power",
            exec_legitimacy == "no constitutionally-defined basis to current regime" ~ 
                "with no contsitutional legitimacy",
            exec_legitimacy == "power constitutionally linked to a single political movement" ~ 
                "constitutionally granted power by single party",
            exec_legitimacy == "presidency independent of legislature; ministry is subject to parliamentary confidence" ~ 
                "independent of and accountable to legislature",
            exec_legitimacy == "presidency is elected by legislature; ministry may be, or not be, subject to parliamentary confidence" ~ 
                "elected by legislature",
            exec_legitimacy == "presidency is independent of legislature" ~ 
                "independent of legislature",
            TRUE ~ '')
        )
    govts$govt_type_short <- paste0(govts$govt, " with ", govts$head_of_state, " head\n", govts$exec_legit_short)
}

```

Because the data came from disparate sources, there were gaps that had to be addressed before proceeding with analysis. For example, many countries had sparser data on literacy rates compared to government spending on education:

```{r data_gap1, ECHO=FALSE}

regions %>%
    select(country_code) %>%
    left_join(edspending, by = c("country_code")) %>%
    group_by(country_code) %>%
    summarize(edspending_year_count = n()) %>%
    full_join(lit, by = c("country_code")) %>%
    group_by(country_code, edspending_year_count) %>%
    summarize(litrate_year_count = n(), .groups = "keep") %>%
    full_join(govts, by = c("country_code")) %>%
    group_by(country_code, edspending_year_count, litrate_year_count) %>%
    summarize(govt_count = n(), .groups = "keep") %>%
    arrange(country_code) %>%
    kable(caption = "<i><font color=#000000><b>Table 1.</b> Observation counts by country</font></i>") %>% 
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 13)

```

In addition, some countries had no country code listed, while others had no literacy rate or educational spending data:

```{r data_gap2, ECHO=FALSE}

regions %>%
    select(country_code) %>%
    left_join(edspending, by = c("country_code")) %>%
    group_by(country_code) %>%
    summarize(edspending_year_count = n()) %>%
    full_join(lit, by = c("country_code")) %>%
    group_by(country_code, edspending_year_count) %>%
    summarize(litrate_year_count = n(), .groups = "keep") %>%
    full_join(govts, by = c("country_code")) %>%
    group_by(country_code, edspending_year_count, litrate_year_count) %>%
    summarize(govt_count = n(), .groups = "keep") %>%
    arrange(country_code) %>%
    filter(is.na(edspending_year_count) | is.na(litrate_year_count) | is.na(govt_count)) %>%
    kable(caption = "<i><font color=#000000><b>Table 2.</b> Data gaps</font></i>") %>% 
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 13)
    
```

For these reasons, the data couldn't be reliably compared on a year-by-year basis. Instead, the data was taken as an aggregate, using mean values for literacy rate and educational spending per country.

```{r data_agg}

##########################
# Data aggregation
##########################

# Calculate mean literacy rate per country
lit_sum <- lit %>% 
    group_by(country, country_code) %>%
    summarize(lit_rate = mean(lit_rate, na.rm = T), .groups = "keep")

# Calculate mean education spending per country
ed_sum <- edspending %>% group_by(country, country_code) %>%
    summarize(pct_gdp = mean(pct_gdp, na.rm = T), .groups = "keep")

# Join the summarized tables
j <- lit_sum %>% 
    full_join(ed_sum, by = c("country_code")) %>%
    full_join(govts, by = c("country_code")) %>%
    left_join(regions, by = c("country_code"))

# Filter out rows with no country code or subregion; these are regional aggregations
j <- j %>%
    filter(str_length(country_code) > 0 & str_length(region) > 0)

# Filter out NaNs
#j <- j %>%
#  filter(!is.na(pct_gdp)) %>%
#  filter(!is.na(lit_rate))

# Select only relevant fields
j <- j %>%
    select(country_code, country = country.x, region, subregion, 
        pct_gdp, lit_rate, govt, head_of_state, exec_legitimacy, exec_legit_short, govt_type_short)

# Add rankings
j$rank_pct_gdp <- rank(desc(j$pct_gdp))
j$rank_lit_rate <- rank(desc(j$lit_rate))

```


### Part 3 - Exploratory data analysis

Data overview:

```{r explore1}

# Data overview
j %>%
    arrange(country_code) %>%
    kable(caption = "<i><font color=#000000><b>Table 3.</b> Data overview</font></i>") %>% 
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 13)

```

Literacy rates (dependent variable):

```{r explore2}

# Summary statistics for literacy rate
describe(j$lit_rate) %>%
    kable(caption = "<i><font color=#000000><b>Table 6.</b> Summary stats - literacy rate</font></i>") %>% 
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 13)

# Histogram of literacy rate
j %>%
    drop_na(lit_rate) %>%
    ggplot() +
    geom_histogram(aes(x = lit_rate), bins = 25, fill = 'lightblue') +
    xlab("Literacy rate") + ylab("Countries") +
    ggtitle("Figure 1. Literacy rate") + 
    theme_light()

```

Educational spending (independent, numerical variable):

```{r explore3}

# Summary statistics for percent GDP education spending
describe(j$pct_gdp) %>%
    kable(caption = "<i><font color=#000000><b>Table 7.</b> Summary stats - educational spending</font></i>") %>% 
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 13)

# Histogram of GDP education spending
j %>%
    drop_na(pct_gdp) %>%
    ggplot() +
    geom_histogram(aes(x = pct_gdp), bins = 20, fill = 'lightblue') +
    xlab("% GDP") + ylab("Countries") +
    ggtitle("Figure 2. Percentage of GDP Spent on Education") +
    theme_light()

```

Regions (independent, categorical variable):

```{r explore4}

# Regions table
regions %>% 
    group_by(region, subregion) %>%
    summarize(n = n(), .groups = "keep") %>%
    select(-n) %>% 
    arrange(region, subregion) %>%
    kable(caption = "<i><font color=#000000><b>Table 4.</b> Regions</font></i>") %>% 
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 13)

# Bar chart of regions
j %>%
    drop_na(subregion) %>%
    group_by(subregion) %>%
    summarize(n = n()) %>%
    ggplot() +
    geom_bar(aes(x = reorder(subregion, n), y = n), fill = 'lightblue', stat = "identity") +
    xlab("Subregion") + ylab("Countries") +
    ggtitle("Figure 3. Subregions") +
    theme_light() +
    coord_flip()

```

Forms of government (independent, categorical variable):

```{r explore5}

# Governments table
govts %>% 
    group_by(govt, head_of_state, exec_legitimacy) %>%
    summarize(num_countries = n(), .groups = "keep") %>%
    arrange(desc(num_countries)) %>%
    kable(caption = "<i><font color=#000000><b>Table 5.</b> Forms of government</font></i>") %>% 
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 13)

# Bar chart of governments
govts %>%
    group_by(govt_type_short) %>%
    summarize(n = n(), .groups = "keep") %>%
    ggplot() +
    geom_bar(aes(x = reorder(govt_type_short, n), y = n), fill = 'lightblue', stat = "identity") +
    xlab("Form of goverment") + ylab("Countries") +
    ggtitle("Figure 3. Forms of government") +
    theme_light() +
    coord_flip()

```

Explore relationships between explanatory and response variables:

```{r explore5}

# Scatter plot of literacy rate vs education spending
j %>% 
    drop_na(pct_gdp, lit_rate) %>%
    ggplot() +
    geom_point(aes(x = pct_gdp, y = lit_rate)) +
    theme_light() +
    xlab("Percent GDP") +
    ylab("Literacy rate") +
    ggtitle("Figure x. Literacy rate vs educational spending")

# Box plots of literacy rate by region
j %>%
    drop_na(lit_rate) %>%
    ggplot() +
    geom_boxplot(aes(x = reorder(subregion, -lit_rate), y = lit_rate)) +
    coord_flip() +
    theme_light() +
    xlab("Subregion") +
    ylab("Literacy rate") +
    ggtitle("Figure x. Literacy rate by subregion")

# Box plots of literacy rate by form of government
j %>%
    drop_na(lit_rate) %>%
    ggplot() +
    geom_boxplot(aes(x = reorder(govt_type_short, -lit_rate), y = lit_rate)) +
    xlab("Form of goverment") + ylab("Literacy rate") +
    ggtitle("Figure x. Literacy rate by form of government") +
    theme_light() +
    coord_flip()

# Set world class for maps
world <- ne_countries(scale = "medium", returnclass = "sf", type = "countries")

# Map of literacy rate
world %>%
    inner_join(j, by = c("iso_a3" = "country_code")) %>%
    drop_na(lit_rate) %>%
    ggplot() +
    geom_sf(aes(fill = lit_rate)) +
    scale_fill_viridis_c(option = "cividis", trans = "sqrt", direction = -1) +
    ggtitle("Figure x. Literacy rate") +
    labs(fill = "Literacy rate")

```

Explore independence of explanatory variables:

```{r explore6}

# Box plots of ed spending by subregion
j %>%
    drop_na(pct_gdp) %>%
    ggplot() +
    geom_boxplot(aes(x = reorder(subregion, -pct_gdp), y = pct_gdp)) +
    coord_flip() +
    theme_light() +
    xlab("Subregion") +
    ylab("% GDP") +
    ggtitle("Figure x. Educational spending by subregion")

# Box plots of ed spending by form of government
j %>%
    drop_na(pct_gdp) %>%
    ggplot() +
    geom_boxplot(aes(x = reorder(govt_type_short, -pct_gdp), y = pct_gdp)) +
    coord_flip() +
    theme_light() +
    xlab("Form of government") +
    ylab("% GDP") +
    ggtitle("Figure x. Educational spending by form of government")

# Bar plot of form of government by region
j %>%
    ggplot() +
    geom_bar(aes(x = region, fill = govt), position = "dodge") +
    coord_flip() +
    theme_light() +
    xlab("Region") +
    ylab("Countries") +
    ggtitle("Figure x. Form of government by subregion")

# Map of subregions
world %>%
    inner_join(j, by = c("iso_a3" = "country_code"))  %>%
    ggplot() +
    geom_sf(aes(fill = subregion.y)) +
    ggtitle("Figure x. Subregions of the world") +
    labs(fill = "Subregion")

# Map of forms of government
world %>%
    inner_join(govts, by = c("iso_a3" = "country_code")) %>%
    ggplot() +
    geom_sf(aes(fill = govt)) +
    scale_fill_brewer(palette="PuBu") +
    ggtitle("Figure x. Forms of government") +
    labs(fill = "Government")

# Map of educational spending
world %>%
    inner_join(j, by = c("iso_a3" = "country_code")) %>%
    drop_na(pct_gdp) %>%
    ggplot() +
    geom_sf(aes(fill = pct_gdp)) +
    scale_fill_viridis_c(option = "cividis", trans = "sqrt", direction = -1) +
    ggtitle("Figure x. Educational spending") +
    labs(fill = "% GDP")

```


### Part 4 - Inference

One goal of the project was to evaluate whether there is a statistically significant difference in literacy rate and educational spending across regions as well as across government types. To evaluate this, a series of ANOVA analyses were performed. It is noted that not all of the conditions for ANOVA were met. Independence was assumed, but while education spending was fairly normal across subregions and government types, the same could not be said of literacy rate, which was strongly left-skewed. Similarly, the box plots also show that variance was not constant for literacy rates, while it was more constant for educational spending.

```{r inf1}

# Literacy rates by subregion
region_summ <- j %>% 
    group_by(region, subregion) %>%
    summarize(
        n_pct_gdp = n(),
        sd_pct_gdp = sd(pct_gdp, na.rm = T),
        mean_pct_gdp = mean(pct_gdp, na.rm = T),
        n_lit_rate = n(),
        sd_lit_rate = sd(lit_rate, na.rm = T),
        mean_lit_rate = mean(lit_rate, na.rm = T),
        .groups = "keep"
    )

# Literacy rates by govt
govt_summ <- j %>% 
    drop_na(govt_type_short) %>%
    group_by(govt_type_short) %>%
    summarize(
        n_pct_gdp = n(),
        sd_pct_gdp = sd(pct_gdp, na.rm = T),
        mean_pct_gdp = mean(pct_gdp, na.rm = T),
        n_lit_rate = n(),
        sd_lit_rate = sd(lit_rate, na.rm = T),
        mean_lit_rate = mean(lit_rate, na.rm = T),
        .groups = "keep"
    ) %>%
    mutate(sd_pct_gdp = ifelse(is.na(sd_pct_gdp), 0, sd_pct_gdp)) %>%
    mutate(sd_lit_rate = ifelse(is.na(sd_lit_rate), 0, sd_lit_rate))

# Find overall rate - ed spending
mean_pct_gdp = mean(j$pct_gdp, na.rm = T)
sd_pct_gdp = sd(j$pct_gdp, na.rm = T)
print(paste0("Mean spending on education across all countries = ", round(mean_pct_gdp, 3), "%, sd = ", round(sd_pct_gdp, 3)))

# Find overall rate - literacy rate
mean_lit_rate = mean(j$lit_rate, na.rm = T)
sd_lit_rate = sd(j$lit_rate, na.rm = T)
print(paste0("Mean literacy rate across all countries = ", round(mean_lit_rate, 3), "%, sd = ", round(sd_lit_rate, 3)))
print("")

# Histograms - ed spending by subregion
j %>%
    drop_na(pct_gdp) %>%
    ggplot(aes(x = pct_gdp)) +
    geom_histogram(binwidth = 0.5) +
    facet_wrap(~subregion, ncol = 3) + 
    xlab("Educational spending (% GDP)") + ylab("")

# Histograms - literacy rate by subregion
j %>%
    drop_na(lit_rate) %>%
    ggplot(aes(x = lit_rate)) +
    geom_histogram(binwidth = 7) +
    facet_wrap(~subregion, ncol = 3) + 
    xlab("Literacy rate (%)") + ylab("")

# Histograms - ed spending by govt
j %>%
    drop_na(pct_gdp) %>%
    ggplot(aes(x = pct_gdp)) +
    geom_histogram(binwidth = 0.5) +
    facet_wrap(~govt_type_short, ncol = 3) + 
    xlab("Educational spending (% GDP)") + ylab("")

# Histograms - literacy rate by govt
j %>%
    drop_na(lit_rate) %>%
    ggplot(aes(x = lit_rate)) +
    geom_histogram(binwidth = 7) +
    facet_wrap(~govt_type_short, ncol = 3) + 
    xlab("Literacy rate (%)") + ylab("")

# ANOVA - ed spending vs subregion
aov_ed_region <- aov(pct_gdp ~ subregion, data = j)
print(paste0("ANOVA results - educatinal spending vs subregion"))
summary(aov_ed_region)

# ANOVA - ed spending vs govt
aov_ed_govt <- aov(pct_gdp ~ govt_type_short, data = j)
print(paste0("ANOVA results - educatinal spending vs government type"))
summary(aov_ed_govt)

# ANOVA - lit rate vs subregion
aov_lit_region <- aov(lit_rate ~ subregion, data = j)
print(paste0("ANOVA results - literacy rate vs subregion"))
summary(aov_lit_region)

# ANOVA - lit rate vs govt
aov_lit_govt <- aov(lit_rate ~ govt_type_short, data = j)
print(paste0("ANOVA results - literacy rate vs government type"))
summary(aov_lit_govt)

```

Next, a linear model was fit to the literacy rate vs ed spending data. It should be noted that not all of the conditions for least squares are *not* met:

1. Linearity: The data appear to be fairly linear, with a positive relationship.
2. Nearly normal residuals: As shown in the histogram and QQ plot, the residuals are significantly skewed to the left.
3. Constant variability: Variability isn't constant, but narrows for higher values.
4. Independent observations: The observations are independent.

A second model was fit that ignored high-leverage points (percent gdp > 7.75%). Contrary to what I expected, the line didn't change much.

```{r lm}

# Linear model - literacy rate vs ed spending
lit_lm <- lm(lit_rate ~ pct_gdp, data = j, na.action = na.omit)
summary(lit_lm)

# Line of best fit
j %>%
    drop_na(pct_gdp, lit_rate) %>%
    ggplot(aes(x = pct_gdp, y = lit_rate)) +
    geom_point() + 
    geom_smooth(formulat = y ~ x, method = "lm", se = T)

# Residuals plot
lit_lm %>%
    ggplot(aes(x = .fitted, y = .resid)) + 
    geom_point() +
    geom_hline(yintercept = 0, linetype = "dashed")

# Residuals histogram
lit_lm %>%
    ggplot(aes(x = .resid)) +
    geom_histogram(, binwidth = 4)

# Probability plot
lit_lm %>%
    ggplot(aes(sample = .resid)) +
    stat_qq()

# Removing high-leverage/influential points
lit_lm_2 <- lm(lit_rate ~ pct_gdp, data = filter(j, pct_gdp < 7.75), na.action = na.omit)
summary(lit_lm_2)
j %>%
    drop_na(pct_gdp, lit_rate) %>%
    filter(pct_gdp < 7.75) %>%
    ggplot(aes(x = pct_gdp, y = lit_rate)) +
    geom_point() + 
    geom_smooth(formulat = y ~ x, method = "lm", se = T)

```

One additional goal of the project was to evaluate whether adding the two categorical variables (subregion and government type) would influence the linear model. First, a backward elimination by p-value strategy was used to eliminate variables. But that resulted in removing the pct_gdp variable and ending up with two categorical variables having p-values greater than 0.05.

```{r multi_p_value}

# Fit linear model
jmod <- j
lm_mult <- lm(lit_rate ~ pct_gdp + subregion + govt_type_short, data = jmod)
summary(lm_mult)
dflm <- data.frame(summary(lm_mult)$coefficients)
colnames(dflm)[4] <- "p_value"
dflm <- dflm %>% arrange(desc(p_value))
new_p <- dflm[[1, 4]]

# Flag to indicate whether pct_gdp was removed from the model or not
pct_gdp_removed <- 0

while(new_p > 0.05 & nrow(dflm) > 1) {
    
    summary(lm_mult)
    # Filter out the variable with the highest p-value
    highest_p <- row.names(dflm)[1]
    print(paste0("Filtering out ", highest_p))
    if (substr(highest_p, 1, 15) == 'govt_type_short') {
        jmod <- jmod %>% filter(govt_type_short != substr(highest_p, 16, str_length(row.names(dflm[1]))))
    } else if (substr(highest_p, 1, 9) == 'subregion') {
        jmod <- jmod %>% filter(subregion != substr(highest_p, 10, str_length(row.names(dflm[1]))))
    } else if (highest_p == 'pct_gdp') {
        pct_gdp_removed <- 1
    }
    
    # Run the model again without the variable
    row_subregion <- nrow(jmod %>% group_by(subregion) %>% summarize(n()))
    row_govt <- nrow(jmod %>% group_by(govt_type_short) %>% summarize(n()))
    if (row_subregion < 2) {
        if (row_govt < 2) {
            if (pct_gdp_removed == 1) {
                # no more variables for the model!
                break
            } else {
                lm_mult <- lm(lit_rate ~ pct_gdp, data = jmod)
            }
        }
        else {
            if (pct_gdp_removed == 1) {
                lm_mult <- lm(lit_rate ~ govt_type_short, data = jmod)
            } else {
                lm_mult <- lm(lit_rate ~ pct_gdp + govt_type_short, data = jmod)
            }
        }
    }
    else {
        if (row_govt < 2) {
            if (pct_gdp_removed == 1) {
                lm_mult <- lm(lit_rate ~ subregion, data = jmod)
            } else {
                lm_mult <- lm(lit_rate ~ pct_gdp + subregion, data = jmod)
            }
        }
        else {
            if (pct_gdp_removed == 1) {
                lm_mult <- lm(lit_rate ~ subregion + govt_type_short, data = jmod)
            } else {
                lm_mult <- lm(lit_rate ~ pct_gdp + subregion + govt_type_short, data = jmod)
            }
        }
    }
    
    # Get the new p-value
    dflm <- data.frame(summary(lm_mult)$coefficients)
    colnames(dflm)[4] <- "p_value"
    dflm <- dflm %>% arrange(desc(p_value))
    new_p <- dflm[[1, 4]]

}

# Print the final model
summary(lm_mult)

```

So a different strategy was employed, using forward selection.

```{r mutli_fwd_sel}

# Fit linear model - pct_gdp
jmod <- j
lm_mult <- lm(lit_rate ~ pct_gdp, data = jmod)
summary(lm_mult)
# adj r-squared = 0.446

# subregion
lm_mult <- lm(lit_rate ~ subregion, data = jmod)
summary(lm_mult)
# adj r-squared = 0.3166

# govt type
lm_mult <- lm(lit_rate ~ govt_type_short, data = jmod)
summary(lm_mult)
# adj r-squared = 0.08609

# Now start with pct_gdp type and add in subregion
lm_mult <- lm(lit_rate ~ pct_gdp + subregion, data = jmod)
summary(lm_mult)
# adj r-squared = 0.3355 (lower than pct_gdp alone)

# pct_gdp type + govt_type
lm_mult <- lm(lit_rate ~ pct_gdp + govt_type_short, data = jmod)
summary(lm_mult)
# adj r-squared = 0.09922 (lower than pct_gdp alone)

# pct_gdp type + subregion + govt_type
lm_mult <- lm(lit_rate ~ pct_gdp + subregion + govt_type_short, data = jmod)
summary(lm_mult)
# adj r-squared = 0.3306 (lower than pct_gdp alone)

```

Using forward selection, the best model is from using educational spending by itself, without subregion or government type as additional predictors.

### Part 5 - Conclusion


### References

$^1$ https://github.com/lukes/ISO-3166-Countries-with-Regional-Codes/blob/master/all/all.csv
$^2$ https://data.worldbank.org/indicator/SE.XPD.TOTL.GD.ZS
$^3$ https://commons.wikimedia.org/wiki/Data:Cross-country_literacy_rates_-_World_Bank,_CIA_World_Factbook,_and_other_sources_(OWID_2762).tab
$^4$ https://en.wikipedia.org/wiki/List_of_countries_by_system_of_government


### Appendix (optional)

Remove this section if you don't have an appendix
